**Effective application of machine learning to bacterial 16S rRNA gene sequencing data**

### 2019-09-10 submission/manuscript.Rmd
changes after Pat meeting 6

---
title: '**Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data**'
csl: mbio.csl
fontsize: 11pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
bibliography: references.bib
---

@@

Begüm D. Topçuoğlu${^1}$, Nicholas A. Lesniak${^1}$, Jenna Wiens${^2}$, Mack Ruffin${^3}$, Patrick D. Schloss^1$\dagger$^

@@

Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce due to the flawed methods that call the validity of developed models into question. Furthermore, the use of black box ML models has hindered the validation of microbial biomarkers. To overcome these challenges, we benchmarked seven different ML models that use fecal 16S rRNA sequences to predict colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases). To show the effect of model selection, we assessed the predictive performance, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost). The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (83.2 h) and hard to interpret. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained much faster (12 min). In this study, we established standards for the development of modeling pipelines for microbiome-associated ML models. Additionally, we showed that ML models should be chosen based on expectations of predictive performance, interpretability and available computational resources.

\newpage
## Importance (needs work)

Prediction of health outcomes using ML is rapidly being adopted by human microbiome studies. However, the developed ML models so far are overoptimistic in terms of validity and predictive performance. Without rigorous ML pipelines, we cannot trust ML models. Before we can speed up progress, we need to slow down, define and implement good ML practices.
\newpage

@@

ML models are useful for understanding the variation in the structure of existing data and to apply that knowledge to make predictions about new data. Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018; @thaiss_persistent_2016; @dadkhah_gut_2019; @flemer_oral_2018; @montassier_pretreatment_2016; @ai_systematic_2017; @dai_multi-cohort_2018; @ai_systematic_2017; @mossotto_classification_2017]. The task of diagnosing an individual with high confidence relies on a ML model that is built with rigorous methods. However, there are common methodological problems across many of these studies that need to be addressed as the field progresses. These include a lack of transparency in which methods are used and how these methods are implemented; developing and evaluating models without a separate held-out test data; large variation between the predictive performance on different folds of cross-validation; and large variation between cross-validation and testing performances. Nevertheless, the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets  [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing analysis frameworks to better use ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019]. More work is needed to further improve reproducibility and minimize over-optimism for model performance.

Among microbiome researchers the lack of transparency in justifying a modelling approach has been due to an implicit assumption that more complex models are better because they are more complex. This has resulted in a trend towards using models such as random forest and neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @wirbel_meta-analysis_2019; @thomas_metagenomic_2019]. Although the more complex models may be better at incorporating non-linear relationships or yield better predictions, they are considered to be black box models because they are not inherently interpretable. These models require post hoc explanations to quantify the importance of each feature in making a prediction and they do not show the structure of how the features are used. Depending on the application of the model, researchers may choose to use different modeling approaches. For example, researchers trying to identify the populations causing a disease would likely want a more interpretable model whereas clinicians may emphasize performance. Although one may feel that they are sacrificing interpretability for performance, that tradeoff may be minimal [@rudin_please_2018; @rudin_optimized_2018]. Regardless, it is important for researchers to articulate why they have selected a specific modelling approach or even compare multiple approaches in the same study.

To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing 7 modeling approaches with the same dataset and pipeline. We built three linear models with different forms of regularization: L2-regularized logistic regression and L1 and L2-regularized support vector machines (SVM) with a linear kernel. We also built four non-linear models: SVM with radial basis function kernel, a decision tree, random forest and XGBoost. We compared the predictive performance, interpretability, and computational efficiency. To demonstrate the performance of these modeling approaches and our pipeline, we used data from a previously published study that sought to classifiy individuals as having normal colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016]. This dataset was selected because it is a relatively large collection of individuals (N=490) connected to a clinically significant disease where there is ample evidence that the disease is driven by variation in the microbiome  [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011]. With this dataset we developed a framework that implements a ML pipeline that can be used for any modeling approach, evaluates predictive performance, and demonstrates how to interpret these models. This framework can be easily applied to other host-associated and environmental microbiome datasets.

@@

Second, we trained seven different models using the training data [Table 1]. We selected models with different classification algorithms and regularization methods. Regularization is a technique that discourages overfitting by penalizing the model for learning the training data too well. For regularized logistic regression and SVM with linear kernel, we used L2 regularization to keep all potentially important features. For comparison, we also trained an L1 regularized SVM model with linear kernel. L1-regularization on microbiome data led to a sparser solution (i.e., force many coefficients to zero). To explore the potential for non-linear relationships among features to improve classification, we trained tree-based models including decision tree, random forest, and XGBoost and we trained an SVM model with non-linear kernel.

Third, fitting of these models require selecting appropriate hyperparameters. Hyperparameters are the rules that are learned from the training set in a classification algorithm. For example, in the linear models the regularization term (C) is a hyperparameter that indicates the penalty for overfitting. Similar to regularization term C, all hyperparameters are tuned to find the best model. We selected hyperparameters by performing 100 five-fold cross-validation (CV) repeats on the training set [Figure 1]. The five-fold CV was also stratified to maintain the overall case and control distribution. We chose the best hyperparameter values for each model based on its CV predictive performance using the area under the receiver operating characteristic curve (AUROC) metric [Figure S1 and S2]. The AUROC ranges from 1.0, where the model perfectly distinguishes between cases and controls, to 0.50, where the model’s predictions are no different from random chance. To select the best performing hyperparameter, we performed a full grid search for hyperparameter settings when training our models. Default hyperparameter settings in previously developed ML packages in R, Python, and Matlab programming languages are inadequate for effective application of classification algorithms and need to be optimized for each new dataset used to generate a model. In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default.

@@

__Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of seven models to classify individuals as having normal colons or SRNs [Figure 2]. The random forest model had significantly higher test AUROC values than the other models for detecting SRNs (Wilcoxon rank sum test, p < 0.01). The median AUROC of the random forest model was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`). L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`) respectively. L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2]. Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularlized logistic regression and linear SVM) performed nearly as well as random forest.

@@

To evaluate the risk for over-optimism of each model, we calculated the range of AUROC values for each model using 100 splits. The range among the testing AUROC values within each model varied by `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 3))` on average across the seven models. If we had only done a single split, then there is a risk that we could gotten lucky or unlucky with the performance of the model. For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` whereas the highest was `r paste(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`. These results showed that depending on the data-split, the testing AUROC values showed great variability [Figure 2]. Therefore, it is important to employ the hierarchical data splits that were included in our pipeline to minimize the risk of over-optimism.

@@

Second, to analyze non-linear models we interpreted the feature importance using permutation importance. Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data. Permutation importance analysis is a posthoc explanation of the model where we randomly permuted non-correlated features individually and groups of perfectly correlated features across the two groups in the held-out test data. We then calculated how much the predictive performance of the model (i.e testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted. We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest [Figure 4]. Among the twenty OTUs with the largest impact for each of these models, there was only one OTU (OTU 822) that was shared among all of the models; however, we found three OTUs (OTU 58, 110, 367) that were important in each of the tree-based models. Similarly, the random forest and XGBoost models, shared four of the most important OTUs (OTU 2, 12, 361, 477). Permutation analysis results also revealed that with the exception of the decision tree model, removal of any individual OTU had a minimal impact on model performance. For example, if OTU 367 were permuted across the diagnoses from the decision tree model, the median AUROC dropped from `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` to `r paste(round(DT$imp, 3))`. In contrast, if we permuted the same OTU from the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(round(RF$imp, 3))`. Effectively, the complexity of the communities was more fully represented in the better performing models [@sze_leveraging_2018; @wirbel_meta-analysis_2019]. At least in this case, it was not possible to distinguish between health and disease using a single OTU. While permutation analysis allowed us to gauge the importance of an OTU, the analysis was post-hoc (i.e. done using the test data) and these results did not allow us to directly interrogate the models to know whether an OTU is associated with classifying a subject as being healthy or having an SRN.

To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S4]. When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S4], 17 out of the 20 top OTUs (e.g. OTU 609, 822, 1239) were deemed important by both approaches. Similarly, for the L2-regularized SVM and L2-regularized logistic regression, 9 and 12 OTUs, respectively, were shared among the two approaches. These results indicate that both approaches are consistent in selecting the most important OTUs.

@@

We highlighted the importance of model interpretation to gain greater biological insights into microbiota-associated diseases. In this study we showcased two different interpretation methods: ranking each OTU by (i) their absolute weights in the trained models and (ii) their impact on the predictive performance based on permutation importance. Human-associated microbial communities have complex correlation structures which create collinearity in the datasets we work with. This can hinder our ability to reliably interpret models because the feature weights of correlated OTUs are influenced by one another [@dormann_collinearity:_2013]. For example if one OTU is highly correlated with another OTU, only one of these OTUs would have a large feature weight thus hiding the importance of its correlated OTU. To capture all important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs. These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments. When we used permutation importance, we took collinearity into consideration by grouping correlated OTUs to determine their impact as a group. We grouped OTUs that had a perfect correlation with each other however, we can reduce the correlation threshold to further investigate the relationships among features. It is important to know the correlation strutures of the data to avoid missinterpreting the models. This is likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronunced due to many genes being correlated with one another because they come from the same chromosome. Therefore, to identify the true underlying microbial factors of a disease, it is crucial to do correlation analyses and further experimentation for biological validation.

In this study, we did not consider all possible modeling approaches. However, the principles highlighted throughout this study apply to all ML modeling tasks with microbiome data. For example, we did not evaluate multicategory classification methods to predict non-binary outcomes. For example, we could have trained models to differentiate between people with normal colons and those with adenomas or carcinomas (k=3 categories). We did not perform this analysis because the clinically relevant diagnosis grouping was between patients with normal colons and those with SRNs. Furthermore, as number of categories to classify increase, more samples are required for each category to train a model. We also did not use regression-based analyses to predict a non-categorical outcome. We have previously used such an approach to train random forest models to predict fecal short-chain fatty acid concentrations based on microbiome data [@sze_fecal_2019]. Our analysis was also limited to shallow learning methods and did not explore deep learning methods such as neural networks. Deep learning methods hold great promise [@geman_deep_2018; @kocheturov_massive_2019; @kim_improved_2018] but microbiome datasets often suffer from having many features and small sample sizes, which makes the deep learning models prone to overfitting. These methods are even more complex than random forest and XGBoost and are considered uninterpretable. There is great potential for applying ML approaches to microbiome data.

Our framework gives structure to investigators wanting to train, evaluate, and interpret their own ML models to identify OTUs that might be biologically relevant. However, deploying microbiome-based models to make clinical diagnoses or predictions is a significantly harder and distinct undertaking. For example, we currently lack standardized methods to collect patient samples, generate sequence data, and report clinical data. We are also challenged by the practical constraints of OTU-based approaches. The de novo algorithms commonly in use are slow, require considerable memory, and result in different OTU assignments as new data are added. Finally, we also need independent validation cohorts to test the performance of a diagnostic model. To realize the potential for using ML approaches with microbiome data, it is necessary we direct our efforts to overcome these challenges.

This study highlighted the need to make educated choices at every step of developing a ML model with microbiome data. We created an aspirational rubric that researchers can use to identify potential pitfalls when using ML in microbiome studies and ways to avoid them [Table 2]. We have highlighted the tradeoffs between model complexity and interpretability, the need for cross-validation to tune hyperparameters, the utility of held-out test sets for evaluating predictive performance, and the importance of considering correlation structures in datasets for reliable interpretation. Furthermore, we underscored the importance of proper experimental design and methods to help us achieve the level of validity and accountability we want from models built for patient health.

@@

\newpage
\includegraphics{Figure_1}
**Figure 1. Machine learning pipeline showing predictive model training and evaluation flowchart.  **  We split the data 80%/20% stratified to maintain the overall label distribution, performed five-fold cross-validation on the training data to select the best hyperparameter setting and then using these hyperparameters to train all of the training data. The model was evaluated on a held-out set of data (not used in selecting the model).
Abbreviations: cvAUROC, cross-validation area under the receiver operating characteristic curve
\newpage
\includegraphics{Figure_2.png}

**Figure 2. Generalization and classification performance of ML models using AUROC values of all cross validation and testing performances. **   The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models. Predictive performance of random forest model was higher than other ML models. The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box. The whiskers show the farthest points that are not outliers. Outliers are data points that are not within 3/2 times the interquartile ranges.
Abbreviations: SRN, screen-relevant neoplasias; AUROC, area under the receiver operating characteristic curve; SVM, support vector machine; XGBoost, extreme gradient boosting
\newpage
\includegraphics[height=17cm, width=12cm]{Figure_3.png}

>>>>

---
title: '**Effective application of machine learning to bacterial 16S rRNA gene sequencing data**'
csl: mbio.csl
fontsize: 11pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
bibliography: references.bib
---

@@

Begüm D. Topçuoğlu${^1}$, Nicholas A. Lesniak${^1}$, Mack Ruffin${^3}$, Jenna Wiens${^2}$, Patrick D. Schloss^1$\dagger$^

@@

Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes, and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but inconsistent adoption of methods call the validity of these models into question. Furthermore, there appears to be a preference by many researchers to favor increased model complexity over interpretability. To overcome these challenges, we trained seven models that used fecal 16S rRNA sequences to predict the presence of colonic screen relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases). We developed a generalizable pipeline to train, validate and interpret the models. To show the effect of model selection, we assessed the predictive performance, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost). The random forest model was best at detecting SRNs with an AUROC of 0.695 but it was slow to train (83.2 h) and hard to interpret. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, it trained much faster (12 min) and was more interpretable. Our analysis showed that ML models should be chosen based on the goal of the study as it will inform our expectations of predictive performance and interpretability.

\newpage
## Importance

Prediction of health outcomes using machine learning (ML) is rapidly being adopted in human microbiome studies. However, these ML models are likely overoptimistic in terms of quantifying predictive performance. Moreover, there is a trend towards using black box models such as random forest and neural networks without a discussion of the difficulty of interpreting such models when trying to identify microbial biomarkers of disease. This work represents a step towards developing better ML practices in microbiome research by implementing a rigorous pipeline and emphasizing the importance of selecting ML models that reflect the goal of the study. These concepts are not particular to study of health outcomes but can also be applied to environmental microbiology studies as well.

\newpage

@@

ML models are useful for understanding the variation in the structure of existing data and to apply that knowledge to make predictions about new data. Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018; @thaiss_persistent_2016; @dadkhah_gut_2019; @flemer_oral_2018; @montassier_pretreatment_2016; @ai_systematic_2017; @dai_multi-cohort_2018; @ai_systematic_2017; @mossotto_classification_2017]. The task of diagnosing an individual with high confidence relies on a model that is built with rigorous methods. However, there are common methodological problems across many of these studies that need to be addressed as the field progresses. These include a lack of transparency in which methods are used and how these methods are implemented; developing and evaluating models without a separate held-out test data; large variation between the predictive performance on different folds of cross-validation; and large variation between cross-validation and testing performances. Nevertheless, the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets  [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing analysis frameworks to better use ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019]. More work is needed to further improve reproducibility and minimize over-optimism for model performance.

Among microbiome researchers, the lack of transparency in justifying a modelling approach has been due to an implicit assumption that more complex models are better because they are more complex. This has resulted in a trend towards using models such as random forest and neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @wirbel_meta-analysis_2019; @thomas_metagenomic_2019]. Although the more complex models may be better at incorporating non-linear relationships or yield better predictions, they are considered to be black box models because they are not inherently interpretable. These models require post hoc explanations to quantify the importance of each feature in making a prediction and they do not show the structure of how the features are used. Depending on the application of the model, researchers may choose to use different modeling approaches. For example, researchers trying to identify the populations causing a disease would likely want a more interpretable model whereas clinicians may emphasize predictive performance. Although one may feel that they are sacrificing interpretability for performance, that tradeoff may be minimal [@rudin_please_2018; @rudin_optimized_2018]. Regardless, it is important for researchers to articulate why they have selected a specific modelling approach or even compare multiple approaches in the same study.

To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing 7 modeling approaches with the same dataset and pipeline. We built three linear models with different forms of regularization: L2-regularized logistic regression and L1 and L2-regularized support vector machines (SVM) with a linear kernel. We also built four non-linear models: SVM with radial basis function kernel, a decision tree, random forest and XGBoost. We compared their predictive performance, interpretability, and computational efficiency. To demonstrate the performance of these modeling approaches and our pipeline, we used data from a previously published study that sought to classifiy individuals as having normal colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016]. This dataset was selected because it is a relatively large collection of individuals (N=490) connected to a clinically significant disease where there is ample evidence that the disease is driven by variation in the microbiome  [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011]. With this dataset we developed a framework that implements a ML pipeline that can be used for any modeling approach, evaluates predictive performance, and demonstrates how to interpret these models. This framework can be easily applied to other host-associated and environmental microbiome datasets.

@@

Second, we trained seven different models using the training data [Table 1]. We selected models with different classification algorithms and regularization methods. Regularization is a technique that discourages overfitting by penalizing the model for learning the training data too well. For regularized logistic regression and SVM with linear kernel, we used L2 regularization to keep all potentially important features. For comparison, we also trained an L1-regularized SVM model with linear kernel. L1-regularization on microbiome data led to a sparser solution (i.e., force many coefficients to zero). To explore the potential for non-linear relationships among features to improve classification, we trained tree-based models including decision tree, random forest, and XGBoost and we trained an SVM model with non-linear kernel.

Third, fitting of these models require selecting appropriate hyperparameters. Hyperparameters are the rules that are learned from the training set in a classification algorithm. For example, in the linear models the regularization term (C) is a hyperparameter that indicates the penalty for overfitting. Similar to regularization term C, all hyperparameters are tuned to find the best model. We selected hyperparameters by performing 100 five-fold cross-validation (CV) repeats on the training set [Figure 1]. The five-fold CV was also stratified to maintain the overall case and control distribution. We chose the best hyperparameter values for each model based on its CV predictive performance using the area under the receiver operating characteristic curve (AUROC) metric [Figure S1 and S2]. The AUROC ranges from 1.0, where the model perfectly distinguishes between cases and controls, to 0.50, where the model’s predictions are no different from random chance. To select the best performing hyperparameter, we performed a full grid search for hyperparameter settings when training our models. Default hyperparameter settings in previously developed ML packages in R, Python, and Matlab programming languages are inadequate for effective application of classification algorithms and need to be optimized for each new ML task. In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default.

@@

__Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of seven models to classify individuals as having normal colons or SRNs [Figure 2]. The random forest model had significantly higher test AUROC values than the other models for detecting SRNs (Wilcoxon rank sum test, p < 0.01). The median AUROC of the random forest model was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`). L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`), respectively. L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2]. Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and SVM with linear kernel) performed nearly as well as random forest.

@@

To evaluate the risk for over-optimism of each model, we calculated the range of AUROC values for each model using 100 data-splits. The range among the testing AUROC values within each model varied by `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 3))` on average across the seven models. If we had only done a single split, then there is a risk that we could gotten lucky or unlucky with the performance of the model. For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` whereas the highest was `r paste(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`. These results showed that depending on the data-split, the testing AUROC values showed great variability [Figure 2]. Therefore, it is important to employ the hierarchical data splits that were included in our pipeline to minimize the risk of over-optimism.

@@

Second, to analyze non-linear models we interpreted the feature importance using permutation importance. Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data. Permutation importance analysis is a posthoc explanation of the model where we randomly permuted non-correlated features individually and groups of perfectly correlated features across the two groups in the held-out test data. We then calculated how much the predictive performance of the model (i.e testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted. We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest [Figure 4]. Among the twenty OTUs with the largest impact, there was only one OTU (OTU 822) that was shared among all of the models; however, we found three OTUs (OTU 58, 110, 367) that were important in each of the tree-based models. Similarly, the random forest and XGBoost models, shared four of the most important OTUs (OTU 2, 12, 361, 477). Permutation analysis results also revealed that with the exception of the decision tree model, removal of any individual OTU had a minimal impact on model performance. For example, if OTU 367 was permuted across the samples in the decision tree model, the median AUROC dropped from `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` to `r paste(round(DT$imp, 3))`. In contrast, if the same OTU was permuted in the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(round(RF$imp, 3))`. Effectively, the complexity of the communities was more fully represented in the better performing models [@sze_leveraging_2018; @wirbel_meta-analysis_2019]. At least in this case, it was not possible to distinguish between health and disease using a single OTU. While permutation analysis allowed us to gauge the importance of an OTU, the analysis was post-hoc (i.e. done using the test data) and these results did not allow us to directly interrogate the models to know whether an OTU is associated with classifying a subject as being healthy or having an SRN.

To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S4]. When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S4], 17 out of the 20 top OTUs (e.g. OTU 609, 822, 1239) were deemed important by both interpretation methods. Similarly, for the L2-regularized SVM and L2-regularized logistic regression, 9 and 12 OTUs, respectively, were shared among the two interpretation methods. These results indicate that both methods are consistent in selecting the most important OTUs.

@@

We highlighted the importance of model interpretation to gain greater biological insights into microbiota-associated diseases. In this study we showcased two different interpretation methods: ranking each OTU by (i) their absolute weights in the trained models and (ii) their impact on the predictive performance based on permutation importance. Human-associated microbial communities have complex correlation structures which create collinearity in the datasets we work with. This can hinder our ability to reliably interpret models because the feature weights of correlated OTUs are influenced by one another [@dormann_collinearity:_2013]. For example if one OTU is highly correlated with another OTU, only one of these OTUs would have a large feature weight thus hiding the importance of its correlated OTU. To capture all important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs. These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments. When we used permutation importance, we took collinearity into consideration by grouping correlated OTUs to determine their impact as a group. We grouped OTUs that had a perfect correlation with each other however, we can reduce the correlation threshold to further investigate the relationships among features. It is important to know the correlation strutures of the data to avoid missinterpreting the models. This is likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronunced due to many genes being correlated with one another because they come from the same chromosome. To identify the true underlying microbial factors of a disease, it is crucial to do correlation analyses and further experimentation for biological validation.

In this study, we did not consider all possible modeling approaches. However, the principles highlighted throughout this study apply to all ML modeling tasks with microbiome data. For example, we did not evaluate multicategory classification methods to predict non-binary outcomes. For example, we could have trained models to differentiate between people with normal colons and those with adenomas or carcinomas (k=3 categories). We did not perform this analysis because the clinically relevant diagnosis grouping was between patients with normal colons and those with SRNs. Furthermore, as number of categories to classify increase, more samples are required for each category to train a model. We also did not use regression-based analyses to predict a non-categorical outcome. We have previously used such an approach to train random forest models to predict fecal short-chain fatty acid concentrations based on microbiome data [@sze_fecal_2019]. Our analysis was also limited to shallow learning methods and did not explore deep learning methods such as neural networks. Deep learning methods hold great promise [@geman_deep_2018; @kocheturov_massive_2019; @kim_improved_2018] but microbiome datasets often suffer from having many features and small sample sizes, which makes the deep learning models prone to overfitting. These methods are even more complex than random forest and XGBoost and are considered uninterpretable. There is great potential for applying ML approaches to microbiome data and we can choose which ML approach is best for the study.

Our framework gives structure to investigators wanting to train, evaluate, and interpret their own ML models to identify OTUs that might be biologically relevant. However, deploying microbiome-based models to make clinical diagnoses or predictions is a significantly harder and distinct undertaking. For example, we currently lack standardized methods to collect patient samples, generate sequence data, and report clinical data. We are also challenged by the practical constraints of OTU-based approaches. The de novo algorithms commonly in use are slow, require considerable memory, and result in different OTU assignments as new data are added. Finally, we also need independent validation cohorts to test the performance of a diagnostic model. To realize the potential for using ML approaches with microbiome data, it is necessary that we direct our efforts to overcome these challenges.

This study highlighted the need to make educated choices at every step of developing a ML model with microbiome data. We created an aspirational rubric that researchers can use to identify potential pitfalls when using ML in microbiome studies and ways to avoid them [Table S1]. We have highlighted the tradeoffs between model complexity and interpretability, the need for cross-validation to tune hyperparameters, the utility of held-out test sets for evaluating predictive performance, and the importance of considering correlation structures in datasets for reliable interpretation. Furthermore, we underscored the importance of proper experimental design and methods to help us achieve the level of validity and accountability we want from models built for patient health.

@@

\newpage
\includegraphics{Figure_1}
**Figure 1. Machine learning pipeline.  **   We split the data to create a training (80%) and held-out test set (20%). The splits were stratified to maintain the overall label distribution. We performed five-fold cross-validation on the training data to select the best hyperparameter setting and then used these hyperparameters to train the models. The model was evaluated on the held-out data set. Abbreviations: cvAUC, cross-validation area under the receiver operating characteristic curve.
\newpage
\includegraphics{Figure_2.png}

**Figure 2. Generalization and classification performance of ML models using AUROC values of all cross validation and testing performances. ** The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models. Predictive performance of random forest model was higher than other ML models. The boxplot shows quartiles at the box ends and the median as the horizontal line in the box. The whiskers show the farthest points that were not outliers. Outliers were defined as those data points that are not within 1.5 times the interquartile ranges. 
\newpage
\includegraphics[height=17cm, width=12cm]{Figure_3.png}
